\chapter{\label{ch:2-litreview}Literature review}

\section{Existing studies - RQ1}
The first research question addresses the causal effect of digital literacy on health. A number of researchers have explored this relationship, particularly over the past decade as concerns about the digital divide have grown considerably \parencite{ariaslopez_digital_2023}. This section will provide an overview of key studies, focusing firstly on the types of health outcomes examined and secondly on the identification strategies employed.

\subsection{Types of health outcomes}
It is widely acknowledged that health encompasses two dimensions: physical health and mental health \parencite{departmentofhealth_relationship_2014,worldbank_secure_2024}. In terms of measurement of health, a further distinction is made between self-rated and objective measures \parencite{baker_what_2004,wu_relationship_2013}. The literature consistently shows that poorer digital literacy is significantly associated with worse outcomes across all three of these dimensions.

Firstly, regarding self-rated health, \textcite{gracia_internet_2009} analysed a survey among older people in Spain, where self-rated health is measured on a five-point Likert scale. They identified a significant association between Internet use and a decreased likelihood of reporting poor health ($OR = 0.41, p = 0.02$), indicating that Internet users generally report better self-rated health than non-users. Similarly, \textcite{falkerhag_internet_2019} analysed cross-sectional data from the Gothenburg H70 Birth Cohort Study in Sweden, using a five-point Likert scale to measure self-rated health. They found a significant correlation between more frequent Internet use and better self-rated health ($\beta = 0.101, p < 0.001$). A similar trend has been observed in Asia; Hong et al. (2017) drew on cross-sectional data from the China Health and Retirement Longitudinal Study (CHARLS). They employed a binary version of self-rated health and discovered that Internet users are more likely to report good health compared to non-users ($OR = 1.727, p < 0.001$).

Secondly, regrading physical health, the current literature is notably scarce. To the best of my knowledge, only two studies have specifically examined the relationship between digital literacy and physical health, focusing on diabetes and lung disease respectively. \textcite{guo_relationships_2021} conducted a cross-sectional survey among 249 diabetes patients across three hospitals in Taiwan, measuring their digital health literacy through a self-developed questionnaire. Their findings indicate that higher digital health literacy has a direct, albeit non-significant, effect on reducing glycated haemoglobin (HbA1c) levels — a lower HbA1c level signifies better diabetes control. \textcite{stellefson_association_2019} carried out a survey among 174 chronic obstructive pulmonary disease (COPD) patients in the US, using the 8-item eHealth Literacy Scale (eHEALS) (Norman \& Skinner, 2006) to assess their digital health literacy. They discovered that patients with higher digital health literacy are significantly less likely to experience symptoms such as chest congestion and tightness, and have significantly better sleep quality.

Thirdly, regarding mental health, \textcite{lee_technology_2018} analysed data from the US National Health and Aging Trends Study, where depression was measured using the Patient Health Questionnaire 2 (PHQ-2) \parencite{kroenke_patient_2003}. They found that greater use of ICTs is significantly associated with a reduced risk of severe depressive symptoms ($OR = 0.28, p < 0.05$). Similarly, \textcite{yang_relationship_2021} conducted a web-based cross-sectional survey in China during the COVID-19 pandemic, assessing respondents' digital health literacy with a self-developed questionnaire. Their findings indicate that higher digital health literacy is inversely associated with the prevalence of several mental health issues, including depression ($r = -0.331$), insomnia ($r = -0.366$) and post-traumatic stress disorder (PTSD) ($r = -0.320$).

In summary, the studies reviewed consistently demonstrate that lower digital literacy is associated with poorer health outcomes. This uniformity suggests a broad consensus in the literature that digital literacy may be a crucial determinant of health among older adults \parencite{ariaslopez_digital_2023,vankessel_digital_2022}. However, whether these findings reflect a valid causal effect requires further investigation. A more detailed discussion on this matter will be provided in Section 2.3.2.

\subsection{Identification strategies}
When investigating the causal effect of digital literacy on health outcomes, most existing studies employ the ``selection on observables" approach, typically within a regression framework, to derive their causal estimates. This method involves controlling for potential confounders that are correlated with both digital literacy and health outcomes. Common control variables frequently used in these analyses include age, gender, marital status, ethnicity, educational attainment, employment status, household income, deprivation and cognitive functioning \parencite{hall_digital_2015,he_factors_2022,mitchell_digital_2019,mcdonald_insights_2004}.

The predominant methodology among these studies is the use of straightforward OLS or logistic regression models that regress health outcomes on digital literacy along with individual-level control variables. The coefficient of digital literacy in these models is then interpreted as the estimated causal effect on health. For instance, when analysing the effect of digital literacy on self-rated health, \textcite{gracia_internet_2009} utilised a logistic regression model when analysing the effect of Internet usage on self-rated health, incorporating controls for age, gender, marital status, and area of residence. 

Some research extends beyond basic models by incorporating more sophisticated methods that control for confounders at multiple levels. For example, \textcite{hong_digital_2017} used a multi-level logistic regression model when assessing the effect of Internet usage on self-rated health, including both individual-level variables and community-level factors such as neighbourhood amenities, healthcare facilities, and community organisations. 

However, it is important to note that relying solely on ``selection on observables" may not yield a valid causal estimate due to the existence for simultaneity bias (i.e., reverse causality), which can significantly skew the relationship between digital literacy and health \parencite{czaja_factors_2013,drentea_association_2008,greer_digital_2019}. A more comprehensive discussion of this limitation will be provided in Section 2.3.2.

\section{Existing studies - RQ2}
The second research question addresses whether increased digitalisation will exacerbate the health inequalities between older adults with low and high digital literacy. Although many researchers advocate that digital health ecosystems should prioritise health equity to prevent worsening existing disparities \parencite{eyrich_bridging_2021,vankessel_digital_2022a}, empirical evidence on this issue remains scarce in the literature. To the best of my knowledge, only one study has delved into this dynamic, employing qualitative interviews to explore how increased digitalisation impacts health inequalities among older adults.

\textcite{alkureishi_digitally_2021} conducted 54 telephone interviews with adult patients at the University of Chicago Medical Center between December 2020 and March 2021, focusing on their experiences with the digital divide and the expansion of ICT-dependent healthcare during the COVID-19 pandemic. Two participants noted that the increasing reliance on ICTs lengthened their wait times for medical visits. They observed that individuals who could engage in virtual visits accessed care more quickly, including faster responses to patient portal messages and more readily available virtual appointments. Additionally, other participants reported that the escalating digitalisation of healthcare services negatively impacted their personal health outcomes due to lack of access and skills required to coordinate their healthcare effectively.

\section{Limitations of existing studies - RQ1}
The findings from the existing literature on the causal effect of digital literacy on health should be evaluated considering the following limitations.

\subsection{Measurement of digital literacy}
As \textcite[p. 21]{vandijk_deepening_2005} proposes, digital literacy comprises two key dimensions: instrumental and informational. The instrumental dimension pertains to hardware literacy, focusing on an individual's ability to operate devices such as mobile phones, tablets, and computers to connect to the Internet. The informational dimension, on the other hand, pertains to software literacy. It focuses on an individual's ability to effectively search, select, process, and utilise information from an abundance of sources and accomplish relevant tasks. Typical activities within this dimension include searching for specific information, engaging in social networking, managing finances, navigating directions, and using government services \parencite{chetty_bridging_2018}.

Measuring digital literacy is a complex challenge, as it encompasses a person's inherent abilities that are not directly quantifiable like educational attainment or as easily observable as gender. In most primary studies where researchers have the capacity to design and administer their own surveys, a standard approach is to use specially crafted and validated questionnaires that assess both the instrumental and informational dimensions of digital literacy. The 8-item eHealth Literacy Scale (eHEALS), developed by Norman and Skinner in 2006, is a widely used instrument that has been extensively validated and is considered highly accurate in measuring digital literacy \parencite{ariaslopez_digital_2023}. An example of this approach can be seen in the study described above by \textcite{stellefson_association_2019}, where they conducted a primary survey among 174 COPD patients and incorporated the eHEALS scale to measure participants' digital literacy.

However, in secondary research where researchers rely on pre-existing datasets, measuring digital literacy becomes particularly challenging if the data itself lacks specifically designed questions for this purpose. Under these circumstances, researchers must devise innovative and credible methods to assess digital literacy effectively, ensuring that their approaches encompass both the instrumental and the informational dimension \parencite{vankessel_digital_2022}. This often requires a careful adaptation of available variables or the development of proxy measures that can accurately reflect the underlying variation in respondents' level of digital literacy.

Most of the existing secondary studies described in Section 2.1 have resorted to using frequency of Internet usage as a proxy measure for digital literacy. However, this approach falls short in capturing the full scope of digital literacy, as it predominantly reflects only the instrumental dimension — merely the ability to access and use the Internet. This measure fails to account for significant variations in the informational dimension, which encompasses the ability to perform a wide array of complex online tasks. For instance, imagine two respondents with the same frequency of Internet usage, one of them uses the Internet solely for chatting with friends whereas the other also uses the Internet for information searching, appointment booking and grocery shopping. It is clear that the latter has a much higher level of digital literacy than the former, but this difference cannot be sufficiently captured simply from their frequency of Internet usage. As a result, the existing studies' approach of using frequency of Internet usage as a proxy for digital literacy lacks sufficient validity, thereby compromising the reliability of findings when applied to analyse the effect of digital literacy on health.

\subsection{Causal inference}
It is crucial to recognise that the statistically significant association between digital literacy and health outcomes observed in many studies does not necessarily imply causation. The causal relationship between digital literacy and health is obscured by two endogeneity issues: omitted variable bias \parencite{hong_digital_2017,lam_exploring_2020} and simultaneity bias \parencite{drentea_association_2008,lam_use_2009}. These biases cast doubt on the observed bivariate associations, indicating they may not represent the genuine causal effect of digital literacy on health. Consequently, without appropriate controls and methodological approaches to address these biases, conclusions drawn from these studies may be misleading.

Omitted variable bias (OVB) occurs when a model fails to include one or more relevant variables that influence both the explanatory variable and the outcome variable \Parencite[p. 59]{angrist_mostly_2009}. As illustrated in Figure \ref{fig:ovb}, if the uncontrolled confounder $Z$ is positively or negatively correlated with both the explanatory variable $X$ and the outcome variable $Y$, the effect of $X$ on $Y$ will be overestimated. Conversely, if $Z$ is positively correlated with one variable and negatively correlated with the other, the effect of $X$ on $Y$ will be underestimated. 

\begin{figure}[h!]
    \centering
    \caption{Omitted variable bias}
    \label{fig:ovb}
    \begin{tikzpicture}[thick]
        \node (X) {X};
        \node (Z) [below right = of X] {Z};
        \node (Y) [above right = of Z] {Y};
          
        \draw[->] (Z) -- (X);
        \draw[->] (Z) -- (Y);
        \draw[->] (X) -- (Y);
      \end{tikzpicture}
\end{figure}

In the context of digital literacy and health, there are a number of widely recognised confounding variables in the literature. These include:

\begin{itemize}
    \item Age: Older age is associated with lower digital literacy and poorer health outcomes \parencite{hall_digital_2015}.
    \item Gender: Males on average have higher digital literacy and poorer health outcomes compared to females \parencite{he_factors_2022}.
    \item Marital status: Married individuals tend to have higher digital literacy and better health outcomes \parencite{he_factors_2022}.
    \item Ethnicity: The effect of ethnicity remains contested. Some suggest that racial disparities contribute to higher digital literacy and improved health outcomes among White individuals compared to non-White individuals \parencite{mitchell_digital_2019}. Conversely, the ``healthy immigrant effect", which involves a selection bias where immigrants may be healthier and more digitally literate than the general population, supports the view that non-White individuals might have higher digital literacy and better health outcomes than their White counterparts \parencite{mcdonald_insights_2004}.
    \item Educational attainment: Higher educational level is associated with higher digital literacy and better health outcomes \parencite{hall_digital_2015}.
    \item Employment status: Employed individuals tend to have higher digital literacy and better health outcomes \parencite{he_factors_2022}. 
    \item Household Income: Higher household income is associated with both higher digital literacy and better health outcomes \parencite{he_factors_2022}.
    \item Deprivation: More deprived individuals tend to have lower digital literacy and poorer health outcomes \parencite{he_factors_2022}.
    \item Cognitive functioning: Better cognitive functioning is associated with both higher digital literacy and better health outcomes \parencite{hong_mediating_2023}.
\end{itemize}

It is important to recognise that the list of confounders provided is not exhaustive. There may be additional observable and/or unobservable confounding variables. In theory, failing to control for even a single variable that influences both digital literacy and health outcomes could introduce omitted variable bias.

Simultaneity bias, also known as reverse causality, occurs when the relationship between variables is bidirectional; that is, not only does the explanatory variable affect the outcome variable, but the outcome variable also influences the explanatory variable \parencite[p. 63]{freedman_specifying_2005}. As illustrated in Figure \ref{fig:sb}, if both variables positively influence each other, the observed association between the explanatory variable $X$ and the outcome variable $Y$ reflects the cumulative effects in both directions. Consequently, this simultaneity bias can lead to an overestimation of the causal effects of $X$ on $Y$. 

\begin{figure}[h!]
    \centering
    \caption{Simultaneity bias}
    \label{fig:sb}
    \begin{tikzpicture}[thick]
        \node (X) {X};
        \node (Y) [right = of X] {Y};
          
        \draw[->] (X) -- (Y);
        \draw[->] (Y) -- (X);
    \end{tikzpicture}
\end{figure}

In the context of digital literacy and health, simultaneity bias is present in two potential scenarios. First, poor health conditions may lead to lower level of digital literacy. Regrading physical health, some diseases can impede an individual's capacity to learn and understand digital skills and knowledge. For instance, it is well-documented that cancer and its treatments often lead to cognitive impairments, affecting memory, processing speed, attention, language, and executive functioning \parencite{giovagnoli_investigation_2012,vonah_impact_2013}. Such cognitive declines can significantly hamper the ability of cancer patients, particularly older adults, to learn digital skills in a world where the Internet still represents a relatively new challenge that demands considerable cognitive engagement. Regarding mental health, research shows that individuals with depression may experience symptoms such as anhedonia, which not only diminishes the desire to engage in new learning experiences but also causes dopaminergic dysfunction in the basal ganglia, further impairing the learning process \parencite{herzallah_depression_2013}. Moreover, conditions like schizophrenia, which can involve hallucinations, might prevent affected individuals from effectively using ICTs and even lead to challenges in recalling how to use such technologies \parencite{greer_digital_2019}.

Secondly, poorer health conditions may paradoxically lead to higher level of digital literacy. This arises because health limitations may restrict individuals' ability to perform everyday tasks in conventional ways (e.g., through face-to-face interactions), compelling them to rely more heavily on the Internet to meet their needs \parencite{domingo_overview_2012}. This increased dependency on digital solutions can motivate them to enhance their digital skills. For example, individuals with mobility issues might find it challenging to do grocery shopping physically and thus might need to develop the digital skills required to order groceries online \parencite{dobransky_disability_2006}. This necessity drives them to become more proficient in digital tools and platforms, thereby inadvertently improving their digital literacy \parencite{drentea_association_2008}. 

Most existing studies, as discussed in Section 2.1.2, adopt a ``selection on observables" approach, which controls for potential confounders to address omitted variable bias. However, this approach does not account for simultaneity bias. Thus, findings from these studies still present a biased estimate of the causal effect of digital literacy on health, underscoring the need for more sophisticated analytical techniques to fully unravel these complex interactions.

\subsection{Range of health outcomes}
As noted in Section 2.1.1, the majority of existing research has primarily concentrated on exploring the impact of digital literacy on self-reported health and mental health outcomes. However, the aspect of physical health, which constitutes a crucial component of an individual's overall health, has received very limited attention. To date, I have only identified two relevant studies on physical health: one for diabetes \parencite{guo_relationships_2021} and one for lung disease (COPD) \parencite{stellefson_association_2019}. This narrow scope overlooks a wide range of common physical health issues, including cardiovascular diseases such as diabetes, as well as non-cardiovascular chronic conditions like cancer. These health problems, given their prevalence among the general population, warrant a more thorough examination to understand their relationship with digital literacy. This gap in the research highlights a critical need for broader investigative efforts to assess the role of digital literacy across a more diverse spectrum of physical health conditions.

\section{Limitations of existing studies - RQ2}
The findings from the existing literature on whether increased digitalisation will exacerbate the health disparities between older adults with low and high digital literacy should be evaluated considering the following limitations. 

\subsection{Lack of quantitative evidence}
As highlighted in Section 2.2, the body of existing research on this question is notably scant, with only one study by \textcite{alkureishi_digitally_2021}, providing qualitative evidence through interviews. Sociology, particularly in public health contexts, is fundamentally a population science \parencite[p. 31]{goldthorpe_sociology_2016}. Goldthorpe emphasises that the probabilistic nature of public health phenomena necessitates quantitative methods to adequately address the variability and heterogeneity within populations and to elucidate patterns and explanations in diverse contexts. Qualitative approaches often faces limitations due to small sample sizes, for instance, the study by \textcite{alkureishi_digitally_2021} only included 54 participants. This small sample size may not adequately represent broader population dynamics and hence cannot reveal genuine patterns \parencite{trafimow_considering_2014}. Consequently, it is imperative to supplement these findings with quantitative research that can leverage larger, more diverse samples to produce more robust and generalisable evidence.

Furthermore, investigating the effects of increased digitalisation on health disparities requires a longitudinal approach, as the term ``increased" implies the need to compare populations at two or more time points. While qualitative longitudinal studies exist and provide valuable insights, they often struggle with consistently comparing data across multiple time points due to the nature of qualitative analysis \parencite{hermanowicz_longitudinal_2013}. In contrast, longitudinal quantitative studies offer a more feasible approach for such analysis. These studies allow for the consistent and quantifiable measurement of changes over time, providing stronger, more reliable estimates that can be effectively compared across different time points \parencite{trafimow_considering_2014}. 

\subsection{Causal inference}
The nature of the second research question - comparing health outcomes between two groups at two time points - intuitively warrants a Difference-in-Difference (DiD) design. DiD involves firstly taking the difference between time points and secondly taking the difference between groups \parencite[p. 229]{angrist_mostly_2009}. Consider a longitudinal survey conducted among older adults: at Wave 1, respondents are categorised into two groups based on their digital literacy levels, and information on their health conditions is collected. Assume that the level of digitalisation increases between Wave 1 and Wave 2. At Wave 2, respondents are revisited and their health conditions are collected again. However, simply applying a DiD analysis based on this group assignment may not yield an accurate estimate of how increased digitalisation impacts the health disparities. This is due to two serious challenges.

First, there may be contamination of group assignment caused by the fluid nature of digital literacy. Individuals' digital literacy level is a not a static attribute and can change over time\parencite{chetty_bridging_2018}. Respondents initially categorised as having low digital literacy in Wave 1 might possess high digital literacy by Wave 2 if reassessed. Consequently, a simple DiD analysis will introduce substantial bias due to this instability in group assignment, thereby distorting the true effect of increased digitalisation on the health disparities between older adults with low and high digital literacy.

Second, the presence of inherent heterogeneity between the two groups can lead to violation of the parallel trends assumption. The parallel trends assumption is an essential assumption in DiD analysis, which posits that, in the absence of the treatment, the treatment and control groups would have followed parallel paths over time \parencite[p. 230]{angrist_mostly_2009}. In this context, it implies that, without an increase in digitalisation, older adults with low and high digital literacy would have experienced similar health trends. However, this assumption may not hold in this case because individuals with varying levels of digital literacy often differ significantly in socio-economic characteristics, which can influence their health trends independently of any changes in digitalisation. For example, digital literacy is often correlated with educational attainment, as those with lower digital literacy are more likely to have lower educational levels. Meanwhile, lower educational levels are also linked to poorer health outcomes (Hall et al., 2015). Thus, without any changes in digitalisation, the group with lower digital literacy could still experience a more rapid health decline compared to their higher digital literacy counterparts. This discrepancy indicates that a simple DiD estimate may not accurately reflect the true impact of increased digitalisation on health disparities. 
